Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2018-01-14 docs/manuscript_plain.tex
preparing manuscript for submission

\documentclass[article]{jss}

@@

%% recommended packages
\usepackage{thumbpdf,lmodern,amssymb,rotating}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Joel L. Pick\\University of New South Wales
   \And Shinichi Nakagawa\\University of New South Wales
   \AND Daniel W.A. Noble\\University of New South Wales
   }
\Plainauthor{Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package}
\Plaintitle{Reproducible, flexible and high throughput data extraction from primary literature: The metaDigitise R package}
\Shorttitle{Data extraction with \pkg{metaDigitise}}

%% - \Abstract{} almost as usual
\Abstract{
Research synthesis, especially in the form of meta-analysis, requires data extraction from primary studies. Meta-analysis synthesizes effect sizes, often calculated from summary statistics of studies. However, exact values of such statistics are commonly hidden in figures. The R package \pkg{metaDigitise} extracts descriptive statistics such as means, standard deviations and, if applicable, correlations from the four types of plots: 1) mean and error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. The package interactively guides the user through data extraction process. Notably, it enables a large-scale extraction using image files, letting the user stop processing, edit and add to the resulting data fame at any point. Further, it facilitates reproducible data extraction from plots with little inter-observer bias, thus, allowing a group of people to participate the extraction of data collaboratively.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{meta-analysis, comparative analysis, data extraction, \proglang{R}, reproducibility, figures, images, summary statistics}
\Plainkeywords{meta-analysis, comparative analysis, data extraction, R, reproducibility, figures, images, summary statistics}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble\\
  Ecology and Evolution Research Centre\\
  School of Biological, Earth and Environmental Sciences\\
  University of New South Wales\\
  Kensington, NSW 2052\\
  Sydney, AUSTRALIA\\
  E-mail: \email{joel.l.pick@gmail.com}\\
  URL: \url{https://github.com/daniel1noble/metaDigitise}
}

@@

@@

\section{Introduction}

@@

\section{Directory Structure, Image Processing and Reproducibility}

@@

\section{Diverse Plot Types}
\pkg{metaDigitise} recognises four main types of plot; Mean and error plots, box plots, scatter plots and histograms, shown in Figure \ref{fig:plot_type}. Each of these can be processed together and integrated into a single output. Alternatively, users can keep like figures together and process them separately.

@@

\subsection{Mean and error plots} 
\fct{metaDigitise} prompts the user to enter group names and allows the user to enter sample sizes ($n$), which are used in downstream processing. The user is then prompted to click on an error bar followed by the mean. Error bars above or below the mean can be clicked - sometimes one is clearer than the other. \fct{metaDigitise} assumes that the error bars are symmetrical. Where the user has clicked the error is displayed in a different colour to the mean (Figure \ref{fig:all_extract}A). The user can subsequently add more groups, edit groups or remove groups. Finally the user is asked what type of error was used in the figure: standard deviation (SD, $\sigma$), standard error (SE) or 95\% confidence intervals (CI95). Standard deviation is calculated from standard error as
\begin{equation}
\sigma = SE \sqrt{n}
\end{equation}
and from 95\% confidence intervals as
\begin{equation}
\sigma = \frac{CI}{1.96} \sqrt{n}
\end{equation}
If the user does not enter a sample size at the time of data extraction (if, for example, the information is not readily available) the SD is not calculated. This can be entered at a later time, however (see below). A function, \fct{error\_to\_sd}, that converts the different error types to SD is also available in the package.

\subsection{Box plots}
As with mean and error plots, \fct{metaDigitise} prompts the user to enter group names and allows the user to enter sample sizes ($n$), which are used in downstream processing. The user is then prompted to click on the maximum ($b$), upper quartile ($q_3$), median ($m$), lower quartile ($q_1$) and minimum ($a$). \fct{metaDigitise} will check that the maximum is greater than the minimum, and return a warning if that is not the case. The user can subsequently add, edit or remove groups. From the extracted data, the mean ($\mu$) and SD are calculated as

@@

\subsection{Scatter plots}
\fct{metaDigitise} prompts the user to enter groups names and then to click on points. Points added by mistake can be deleted. The user can subsequently add groups, edit groups (add or remove points) or delete groups. Different groups are plotted in different colours and shapes, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}C). Mean, SD and sample size are calculated from the clicked points, for each group. Where the sample size from the clicked points does not match a known sample size (e.g. if there are overlaid points), the user can enter an alternate sample size.

\subsection{Histograms}
\fct{metaDigitise} prompts the user to click on the top corners of each bar. Bars can subsequently be deleted. 
For each bar a midpoint (m; mean x coordinates) and a frequency (f; mean y coordinates, rounded to the nearest integer) is calculated. The sample size, mean and SD are calculated as:
\begin{equation}
n = \sum_{i=1}^n{f_i}
\end{equation}

@@

\section{Extracting Data From Plots}
%% scatterplot from airquality data

@@

\subsection{Adding new figures}

@@

\section{Re-importing, Editing and Plotting Previously Digitised data}

@@

\section{Additional Features}

\subsection{Figure Rotation and Adjustment}
Figures may have been extracted from old publications, for example from scanned images, and so are not perfectly orientated on the image. This will make the calibration of the points in the figure from the image problematic. \fct{metaDigitise} allows users to rotate the image. By clicking two points on the x-axis, metaDigitse calculates the angle needed to rotate the image so the x-axis is horizontal, and rotates it. (Figure \ref{fig:rotate}A,B)

@@

\subsection{Obtaining Processed Data}

@@

\subsection{Adding sample sizes to previous Digitisations}
%Including addition of N later
In many cases important information, such as sample sizes, may not be readily available or clear when digitising figures. In these circumstances users will have answered `no' to the question about whether they have sample sizes or not while digitising. To expedite finding and adding in these sample sizes to do the necessary calculations (if for example a figure presented 95\% CI's or standard errors), \fct{metaDigitise} has s specific edit option that allows users to enter in previously omitted sample sizes. It works by first identifying the missing sample sizes in the digitised output, re-plotting the relevant figure and then prompting the user to enter the sample sizes for the relevant groups in the figure, one by one. As an example, assume that we were missing sample sizes for two groups in 002\_Doe\_2013\_Fig1.png:

@@

\section{Inter-observer Variability and Validation}

\subsection{Inter-observer variability in digitisations}
In order to evaluate the consistency of digitisation using \pkg{metaDigitise} between users, we simulated a dataset of two traits with two different groups. These data were then used to construct plots of the four different types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice for each given plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. 15 independent digitisers were provided with a directory with all 14 figures in a randomised order. Digitisers ran \pkg{metaDigistise} on their own computers, across different operating systems (including Mac, Windows and Linux). Digitisers varied in their level of experience, from people with experience of meta-analyses or comparative work to those without any science background. We asked users to digitise all 14 figures and collected the mean, standard deviation and correlation coefficient (for scatterplots) generated by \fct{metaDigitise} for every plot digitised. We transformed these data to standardized differences as

@@

\subsection{Testing the accuracy of digitisations}
To test how accurate \pkg{metaDigitise} is at matching points to their true values, we generated four random scatterplots, each with 20 data points, and digitised these with \fct{metaDigitise}. This was done by one digitiser, as there is no detectable between user variation. Data digitised using \pkg{metaDigitise} was essentially perfectly correlated with the true simulated data for both the \textit{x}-variable (Pearson's correlation; \textit{r} = 0.9999915, \textit{t} = 2137.4, \textit{df} = 78, \textit{p} < 0.001) and \textit{y}-variable (\textit{r} = 0.9999892, \textit{t} = 1897.8, \textit{df} = 78, \textit{p} < 0.001).

\section{Discussion and Conclusions}

@@

%While \pkg{metaDigitise} extracts descriptve statistics from figures, given these are most often presented in figures and are usually the most robust sources of information for calculating effect size statistics \citep{Noble2017}, users may also like to compare estimates from inferential statistics from those derived with descriptive statistics from a paper. Comparing effects sizes derived from both sources can be useful in identifying uncertainties and problems within a paper. In the future, we hope to provide functions to easily convert inferential statistics to standardised effect size estimates, which can seamlessly be integrated with summary statistics from \pkg{metaDigitise}, to calculate equivalent standardised effect size estimates and their sampling variance.

@@

\bibliography{metaDigitise}

>>>>

\documentclass[12pt]{article}

@@

%% recommended packages
\usepackage{amssymb,rotating,natbib,graphicx,fancyvrb,fullpage} 
\usepackage[parfill]{parskip} 
\usepackage{geometry}            % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper} 
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{lineno}                                                                            
\doublespacing
\linenumbers
\renewcommand\linenumberfont{\normalfont\small}

%% new custom commands
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\class}[1]{`\code{#1}'}
%\newcommand{\fct}[1]{#1}
\newcommand{\fct}[1]{\texttt{#1()}}
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
\let\proglang=\textsf

@@

\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

\raggedright

\textbf{Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package}

Joel L. Pick$^{1,*}$, Shinichi Nakagawa$^1$, Daniel W.A. Noble$^1$

$^1$
  Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences,  University of New South Wales, Kensington, NSW 2052, Sydney, AUSTRALIA

 $^*$Corresponding Author: joel.l.pick@gmail.com\\

%% - \Abstract{} almost as usual
\section*{Abstract}
Research synthesis, especially in the form of meta-analysis, requires data extraction from primary studies. Meta-analysis synthesizes effect sizes, often calculated from summary statistics of studies. However, exact values of such statistics are commonly hidden in figures. The R package \pkg{metaDigitise} extracts descriptive statistics such as means, standard deviations and, if applicable, correlations from the four types of plots: 1) mean and error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. The package interactively guides the user through data extraction process. Notably, it enables a large-scale extraction using image files, letting the user stop processing, edit and add to the resulting data fame at any point. Further, it facilitates reproducible data extraction from plots with little inter-observer bias, thus, allowing a group of people to participate the extraction of data collaboratively.

Keywords: meta-analysis, comparative analysis, data extraction, \proglang{R}, reproducibility, figures, images, summary statistics

@@

\section*{Introduction}

@@

\section*{Directory Structure, Image Processing and Reproducibility}

@@

\section*{Diverse Plot Types}
\pkg{metaDigitise} recognises four main types of plot; Mean and error plots, box plots, scatter plots and histograms, shown in Figure \ref{fig:plot_type}. Each of these can be processed together and integrated into a single output. Alternatively, users can keep like figures together and process them separately.

@@

\subsection*{Mean and error plots} 
\fct{metaDigitise} prompts the user to enter group names and allows the user to enter sample sizes ($n$), which are used in downstream processing. The user is then prompted to click on an error bar followed by the mean. Error bars above or below the mean can be clicked - sometimes one is clearer than the other. \fct{metaDigitise} assumes that the error bars are symmetrical. Where the user has clicked the error is displayed in a different colour to the mean (Figure \ref{fig:all_extract}A). The user can subsequently add more groups, edit groups or remove groups. Finally the user is asked what type of error was used in the figure: standard deviation (SD, $\sigma$), standard error (SE) or 95\% confidence intervals (CI95). Standard deviation is calculated from standard error as
\begin{equation}
\sigma = SE \sqrt{n}
\end{equation}
and from 95\% confidence intervals as
\begin{equation}
\sigma = \frac{CI}{1.96} \sqrt{n}
\end{equation}
If the user does not enter a sample size at the time of data extraction (if, for example, the information is not readily available) the SD is not calculated. This can be entered at a later time, however (see below). A function, \fct{error\_to\_sd}, that converts the different error types to SD is also available in the package.

\subsection*{Box plots}
As with mean and error plots, \fct{metaDigitise} prompts the user to enter group names and allows the user to enter sample sizes ($n$), which are used in downstream processing. The user is then prompted to click on the maximum ($b$), upper quartile ($q_3$), median ($m$), lower quartile ($q_1$) and minimum ($a$). \fct{metaDigitise} will check that the maximum is greater than the minimum, and return a warning if that is not the case. The user can subsequently add, edit or remove groups. From the extracted data, the mean ($\mu$) and SD are calculated as

@@

\subsection*{Scatter plots}
\fct{metaDigitise} prompts the user to enter groups names and then to click on points. Points added by mistake can be deleted. The user can subsequently add groups, edit groups (add or remove points) or delete groups. Different groups are plotted in different colours and shapes, with a legend at the bottom of the figure (Figure \ref{fig:all_extract}C). Mean, SD and sample size are calculated from the clicked points, for each group. Where the sample size from the clicked points does not match a known sample size (e.g. if there are overlaid points), the user can enter an alternate sample size.

\subsection*{Histograms}
\fct{metaDigitise} prompts the user to click on the top corners of each bar. Bars can subsequently be deleted. 
For each bar a midpoint (m; mean x coordinates) and a frequency (f; mean y coordinates, rounded to the nearest integer) is calculated. The sample size, mean and SD are calculated as:
\begin{equation}
n = \sum_{i=1}^n{f_i}
\end{equation}

@@

\section*{Extracting Data From Plots}
%% scatterplot from airquality data

@@

\subsection*{Adding new figures}

@@

\section*{Re-importing, Editing and Plotting Previously Digitised data}

@@

\section*{Additional Features}

\subsection*{Figure Rotation and Adjustment}
Figures may have been extracted from old publications, for example from scanned images, and so are not perfectly orientated on the image. This will make the calibration of the points in the figure from the image problematic. \fct{metaDigitise} allows users to rotate the image. By clicking two points on the x-axis, metaDigitse calculates the angle needed to rotate the image so the x-axis is horizontal, and rotates it. (Figure \ref{fig:rotate}A,B)

@@

\subsection*{Obtaining Processed Data}

@@

\subsection*{Adding sample sizes to previous Digitisations}
%Including addition of N later
In many cases important information, such as sample sizes, may not be readily available or clear when digitising figures. In these circumstances users will have answered `no' to the question about whether they have sample sizes or not while digitising. To expedite finding and adding in these sample sizes to do the necessary calculations (if for example a figure presented 95\% CI's or standard errors), \fct{metaDigitise} has s specific edit option that allows users to enter in previously omitted sample sizes. It works by first identifying the missing sample sizes in the digitised output, re-plotting the relevant figure and then prompting the user to enter the sample sizes for the relevant groups in the figure, one by one. As an example, assume that we were missing sample sizes for two groups in 002\_Doe\_2013\_Fig1.png:

@@

\section*{Inter-observer Variability and Validation}

\subsection*{Inter-observer variability in digitisations}
In order to evaluate the consistency of digitisation using \pkg{metaDigitise} between users, we simulated a dataset of two traits with two different groups. These data were then used to construct plots of the four different types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice for each given plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. 14 independent digitisers were provided with a directory with all 14 figures in a randomised order. Digitisers ran \pkg{metaDigistise} on their own computers, across different operating systems (including Mac, Windows and Linux). Digitisers varied in their level of experience, from people with experience of meta-analyses or comparative work to those without any science background. We asked users to digitise all 14 figures and collected the mean, standard deviation and correlation coefficient (for scatterplots) generated by \fct{metaDigitise} for every plot digitised. We transformed these data to standardized differences as

@@

\subsection*{Testing the accuracy of digitisations}
To test how accurate \pkg{metaDigitise} is at matching points to their true values, we generated four random scatterplots, each with 20 data points, and digitised these with \fct{metaDigitise}. This was done by one digitiser, as there is no detectable between user variation. Data digitised using \pkg{metaDigitise} was essentially perfectly correlated with the true simulated data for both the \textit{x}-variable (Pearson's correlation; \textit{r} = 0.9999915, \textit{t} = 2137.4, \textit{df} = 78, \textit{p} < 0.001) and \textit{y}-variable (\textit{r} = 0.9999892, \textit{t} = 1897.8, \textit{df} = 78, \textit{p} < 0.001).

\section*{Discussion and Conclusions}

@@

@@

\bibliographystyle{FuncEcol.bst}
\bibliography{metaDigitise}
